Instructions for developing new pages for the MarineOmics webpage

### 2023-12-14 DRAFT_DNA_methylation.Rmd
moved text to alignment - added code explanation

---
title: 'DNA Methylation Analyses'
subtitle: '*Steven Roberts*, *Sam White*, and *Yaamini Venkataraman*'
bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
---
Initial publication year: TBD <br>
[How to cite](https://marineomics.github.io/#How_to_Cite)

@@

The tutorial will cover the following:
- Sequence quality assessment
- Read alignment
- Methylation quantification
- File conversions
- Potential downstream applications. Although this tutorial will not provide code for downstream applications such as differential methylation analysis, it will link to other open-access resources and examples.

@@

DNA methylation at CpG locus can vary to 0-100%. Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage. Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below. From an taxa perspective it can vary WIDELY on what to expect. The figure below shows the mehthylation ratio of CpGs across the oyster genome. In some marine invertebrates we see ~15% methylation of CpGs (defined as > 75% methylated) across the genome with most CpGs unmethylated. In these taxa methylation usually occurs in gene bodes.

```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
```

@@

    /gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
    --output_dir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD \
    --paired \
    --fastqc_args \
    --outdir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD/FASTQC \
    --threads 28 \
    --illumina \
    --clip_R1 10 \
    --clip_R2 10 \
    --three_prime_clip_R1 10 \
    --three_prime_clip_R2 10 \
    --path_to_cutadapt /gscratch/srlab/programs/miniconda3/bin/cutadapt \
    /gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \

@@

@@

    # Directories and programs
    bismark_dir="/programs/Bismark-0.21.0"
    bowtie2_dir="/programs/bowtie2-2.3.4.1-linux-x86_64/"
    genome_folder="/where/the/fastafile/lives/"

    ${bismark_dir}/bismark_genome_preparation \
    --verbose \
    --parallel 28 \
    --path_to_aligner ${bowtie2_dir} \
    ${genome_folder}

You should expect to prepared genome with new directories including

    ./Bisulfite_Genome
    ./Bisulfite_Genome/GA_conversion
    ./Bisulfite_Genome/CT_conversion

@@

    find ${reads_dir}*_R1_001_val_1.fq.gz \
    | xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
    --path_to_bowtie ${bowtie2_dir} \
    -genome ${genome_folder} \
    -p 4 \
    -score_min L,0,-0.6 \
    --non_directional \
    -1 ${reads_dir}{}_R1_001_val_1.fq.gz \
    -2 ${reads_dir}{}_R2_001_val_2.fq.gz \
    -o Mcap_tg

@@

    find *.bam | \
    xargs basename -s .bam | \
    xargs -I{} ${bismark_dir}/deduplicate_bismark \
    --bam \
    --paired \
    {}.bam

@@

    ${bismark_dir}/bismark_methylation_extractor \
    --bedGraph \
    --counts \
    --comprehensive \
    --merge_non_CpG \
    --multicore 28 \
    --buffer_size 75% \
    *deduplicated.bam

@@

    NC_035784.1 141 141 37.5    3   5
    NC_035784.1 142 142 100 2   0
    NC_035784.1 155 155 70  7   3
    NC_035784.1 156 156 100 2   0
    NC_035784.1 291 291 0   0   2
    NC_035784.1 292 292 0   0   3
    NC_035784.1 313 313 0   0   1
    NC_035784.1 314 314 66.6666666666667    2   1
    NC_035784.1 470 470 66.6666666666667    4   2
    NC_035784.1 611 611 0   0   4

@@

    <chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>

@@

    find *deduplicated.bismark.cov.gz \
    | xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz \
    | xargs -I{} ${bismark_dir}/coverage2cytosine \
    --genome_folder ${genome_folder} \
    -o {} \
    --merge_CpG \
    --zero_based \
    {}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz

@@

    NC_035785.1 217 219 100.000000  17  0
    NC_035785.1 523 525 87.500000   7   1
    NC_035785.1 556 558 50.000000   5   5
    NC_035785.1 727 729 100.000000  16  0
    NC_035785.1 1330    1332    0.000000    0   2
    NC_035785.1 1403    1405    0.000000    0   2
    NC_035785.1 1494    1496    66.666667   2   1
    NC_035785.1 1747    1749    100.000000  8   0
    NC_035785.1 2024    2026    100.000000  24  0
    NC_035785.1 2054    2056    93.333333   14  1

@@

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
      > "${STEM}"_10x.bedgraph
    done

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4}}' \
      > "${STEM}"_5x.bedgraph
    done

@@

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
      > "${STEM}"_10x.tab
    done

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
      > "${STEM}"_5x.tab
    done

@@

    # Sort files for methylkit and IGV
    find *.bam | \
    xargs basename -s .bam | \
    xargs -I{} ${samtools} \
    sort --threads 28 {}.bam \
    -o {}.sorted.bam

    # Index sorted files for IGV
    find *.sorted.bam | \
    xargs basename -s .sorted.bam | \
    xargs -I{} ${samtools} \
    index -@ 28 {}.sorted.bam

>>>>

---
title: 'DNA Methylation Analyses'
subtitle: '*Steven Roberts*, *Sam White*, and *Yaamini Venkataraman*'
bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
---

Initial publication year: TBD <br> [How to cite](https://marineomics.github.io/#How_to_Cite)

@@

The tutorial will cover the following: - Sequence quality assessment\
- Read alignment\
- Methylation quantification\
- File conversions\
- Potential downstream applications. Although this tutorial will not provide code for downstream applications such as differential methylation analysis, it will link to other open-access resources and examples.

@@

*Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue.* DOI: [10.7717/peerj.215/fig-1](https://doi.org/10.7717/peerj.215/fig-1)

@@

```         
/gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
--output_dir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD \
--paired \
--fastqc_args \
--outdir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD/FASTQC \
--threads 28 \
--illumina \
--clip_R1 10 \
--clip_R2 10 \
--three_prime_clip_R1 10 \
--three_prime_clip_R2 10 \
--path_to_cutadapt /gscratch/srlab/programs/miniconda3/bin/cutadapt \
/gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \
```

@@

DNA methylation at CpG locus can vary to 0-100%. Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage. Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below. From an taxa perspective it can vary WIDELY on what to expect. The figure below shows the mehthylation ratio of CpGs across the oyster genome. In some marine invertebrates we see \~15% methylation of CpGs (defined as \> 75% methylated) across the genome with most CpGs unmethylated. In these taxa methylation usually occurs in gene bodies.

```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center"}
knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
```

@@

```         
# Directories and programs
bismark_dir="/programs/Bismark-0.21.0"
bowtie2_dir="/programs/bowtie2-2.3.4.1-linux-x86_64/"
genome_folder="/where/the/fastafile/lives/"

${bismark_dir}/bismark_genome_preparation \
--verbose \
--parallel 28 \
--path_to_aligner ${bowtie2_dir} \
${genome_folder}
```

1.  **Bismark Genome Preparation Command**:

    -   **`${bismark_dir}/bismark_genome_preparation`**: This is the command to run the genome preparation part of Bismark. The **`${bismark_dir}`** variable is expanded to the path where Bismark is installed, so the script knows where to find the **`bismark_genome_preparation`** program.

    -   **`--verbose`**: This flag makes the program output more detailed information about what it is doing, which is helpful for debugging or understanding the process.

    -   **`--parallel 28`**: This option tells Bismark to use 28 threads in parallel to speed up the process.

    -   **`--path_to_aligner ${bowtie2_dir}`**: This specifies the path to the aligner (Bowtie2) that Bismark will use. The **`${bowtie2_dir}`** variable is expanded to the path where Bowtie2 is installed.

    -   **`${genome_folder}`**: Finally, this specifies the location of the genome files. The script uses the **`${genome_folder}`** variable, which holds the path to these files.

This will result in a prepared genome with associated new directories including

```         
./Bisulfite_Genome
./Bisulfite_Genome/GA_conversion
./Bisulfite_Genome/CT_conversion
```

@@

```         
find ${reads_dir}*_R1_001_val_1.fq.gz \
| xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
--path_to_bowtie ${bowtie2_dir} \
-genome ${genome_folder} \
-p 4 \
-score_min L,0,-0.6 \
--non_directional \
-1 ${reads_dir}{}_R1_001_val_1.fq.gz \
-2 ${reads_dir}{}_R2_001_val_2.fq.gz \
-o Mcap_tg
```

@@

```         
find *.bam | \
xargs basename -s .bam | \
xargs -I{} ${bismark_dir}/deduplicate_bismark \
--bam \
--paired \
{}.bam
```

@@

```         
${bismark_dir}/bismark_methylation_extractor \
--bedGraph \
--counts \
--comprehensive \
--merge_non_CpG \
--multicore 28 \
--buffer_size 75% \
*deduplicated.bam
```

@@

```         
NC_035784.1 141 141 37.5    3   5
NC_035784.1 142 142 100 2   0
NC_035784.1 155 155 70  7   3
NC_035784.1 156 156 100 2   0
NC_035784.1 291 291 0   0   2
NC_035784.1 292 292 0   0   3
NC_035784.1 313 313 0   0   1
NC_035784.1 314 314 66.6666666666667    2   1
NC_035784.1 470 470 66.6666666666667    4   2
NC_035784.1 611 611 0   0   4
```

@@

```         
<chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
```

@@

```         
find *deduplicated.bismark.cov.gz \
| xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz \
| xargs -I{} ${bismark_dir}/coverage2cytosine \
--genome_folder ${genome_folder} \
-o {} \
--merge_CpG \
--zero_based \
{}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz
```

@@

```         
NC_035785.1 217 219 100.000000  17  0
NC_035785.1 523 525 87.500000   7   1
NC_035785.1 556 558 50.000000   5   5
NC_035785.1 727 729 100.000000  16  0
NC_035785.1 1330    1332    0.000000    0   2
NC_035785.1 1403    1405    0.000000    0   2
NC_035785.1 1494    1496    66.666667   2   1
NC_035785.1 1747    1749    100.000000  8   0
NC_035785.1 2024    2026    100.000000  24  0
NC_035785.1 2054    2056    93.333333   14  1
```

@@

```         
for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
  > "${STEM}"_10x.bedgraph
done

for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4}}' \
  > "${STEM}"_5x.bedgraph
done
```

@@

```         
for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
  > "${STEM}"_10x.tab
done

for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
  > "${STEM}"_5x.tab
done
```

@@

```         
# Sort files for methylkit and IGV
find *.bam | \
xargs basename -s .bam | \
xargs -I{} ${samtools} \
sort --threads 28 {}.bam \
-o {}.sorted.bam

# Index sorted files for IGV
find *.sorted.bam | \
xargs basename -s .sorted.bam | \
xargs -I{} ${samtools} \
index -@ 28 {}.sorted.bam
```
